---
title: "tidy-model"
author: "Jonathan Sebastiani"
date: "2025-02-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('tidyverse') ; library('modeldata')
``` 

## What is that sound?

In this report, we examine what factors predict a cricket's chirp rate.

```{r summary, echo = FALSE}
summary(crickets)
ggplot(crickets, aes(x = rate)) +
  geom_histogram(bins = 8) + 
  ggtitle("Distribution of the chirp rate of crickets") +
  xlab('Chirp rate (per min.)')
```

## Temperature affects chirp rate

```{r temp, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')

temp_lm <- lm(rate ~ temp, crickets)

summary.lm(temp_lm)
```

Based on a scatter plot of temperature and chirping and a correlation test, it seems that as temperature increases one degree, the rate of chirping increases about 4.2 chirps per minute.

## Species-specific effects of temperature on chirping

```{r species, echo= FALSE}
ggplot(crickets, aes(x = temp, y = rate, color = species)) +
  geom_point() + 
  geom_smooth(method = 'lm') +
  ggtitle("Plot of temperature and chirp rate for two species of crickets") +
  ylab('Chirp rate (per min.)') +
  xlab('Temperature (Celsius)')
  
  
species_lm <- lm(rate ~ temp + species, crickets)

summary.lm(species_lm)
```

Based on the scatter plot of the two species, we can see that there are a high correlation between temperature and chirp rate. When we clump the species together, the linear model explains about 92% of the variance, but when we separate into two different species, the model explains 99% of the variance in the data! This is a much better model.